---
title: "Maxlik and Poisson"
output: html_document
date: 'Summer Semester 2022'
author: 'Cyryl Leszczyński'
---
Assume that $X$ follows Poisson distribution given by $P(X=x, \lambda_i) = \frac{\lambda^x e^{-\lambda}}{x!}$
where $\lambda_i = \theta_0 + \theta_1 \times z_i$, $\theta_0=0.5$, $\theta_1=0.5$, and $z_i \sim \text{Bern}(0.7)$ and number of observations is equal to $n=10,000$. \
        1. generate $z_i$, \
        2. generate $\lambda_i$ according to $\theta_0 + \theta_1 \times z_i$, \
        3. generate $X \sim Poisson(\lambda_i)$ \
        4. derive log-likelihood, gradient and hessian (by hand, \LaTeX code for equations are preferred), \
        5. obtain MLE of $\boldsymbol{\theta} = (\theta_0, \theta_1)$ using Newton-Raphson method (gradient and hessian should be applied). \
        
---

### Step 1: Generating $z_i$ (and setting up variables)

```{r}
set.seed(0)
n <- 10000
theta_0 <- 0.5
theta_1 <- 0.5

z_i <- rbinom(n, 1, 0.7) # Bernoulli distribution is the same as binomial with size = 1
```

### Step 2: generate $\lambda_i$ according to $\theta_0 + \theta_1 \times z_i$

```{r}
lambda_i <- theta_0 + (theta_1 * z_i)
```

### Step 3: generate $X \sim Poisson(\lambda_i)$
```{r}
X <- rpois(n,lambda_i)
```

### Step 4: derive log-likelihood, gradient and hessian
As defined by the task, the distribution is given by $P(X=x, \lambda_i) = \frac{\lambda^x e^{-\lambda}}{x!}$. \
The log-likelihood of this function is $log(\frac{\lambda^x e^{-\lambda}}{x!}) = -n\cdotp\lambda-\sum log(x!)+ log(\lambda)\cdotp\sum x$.\
Gradient of the above function is its first derivative: $\frac{\sum x}{\lambda} - n$.
The Hessian is its second: $-\frac{\sum x}{\lambda ^ 2}$

### Step 5: obtain MLE of $\boldsymbol{\theta} = (\theta_0, \theta_1)$ using Newton-Raphson method (gradient and hessian should be applied).
Log-likelihood function:
```{r}
loglikelihood <- function(par, x) {
  ll <- -length(x)*par - sum(log(factorial(x))) + log(par) * sum(x)
  ll
}
```
Gradient of the log-likelihood:
```{r}
gradient <- function(par, x)  {
  grad <- sum(x) / par - length(x)
  grad
}
```
Hessian of the log-likelihood:
```{r}
hessian <- function(par, x) {
  hess <- -(sum(x)/par^2)
  hess
}
```
MLE using the Newton-Rhapsod method:
```{r}
install.packages("maxLik")
library("maxLik")

mle_final <- maxLik(logLik = loglikelihood, grad = gradient, hess = hessian, start = 0.5, x = X, method = "NR")
summary(mle_final)
mle_final
```